Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-02-28 manuscript.tex
Tied the intro together. Need proof-reading

Satellite missions dedicated to measuring the Earths gravity field
(like CHAMP, GRACE, and GOCE)
have provided geophysicists with almost uniform and global data coverage.

These new data have enabled interpretations on regional and global scales
\citep[e.g.][]{Reguzzoni2013,Braitenberg2015}.

Modeling at such scales requires taking into account the curvature of the
Earth and calculating gravity gradients as well as the traditional
gravitational acceleration.

A common approach to achieve this is
to discretize the Earth into tesseroids (spherical prisms)
instead of rectangular prisms.

However, the integral formula for the gravitational effects of a tesseroid
must be solved numerically.

Approaches to this numerical integration include
Taylor series expansion \citep{Heck2007, Grombein2013}
and the Gauss-Legendre Quadrature \citep{Asgharzadeh2007}.

Taylor series expansion produces accurate results at low latitudes but
presents a decrease in accuracy towards the polar regions.

This is attributed to tesseroids degenerating into an approximately triangular
shape at the poles.

The Gauss-Legendre Quadrature (GLQ) integration consists of approximating
the volume integral by a weighted sum of the effect of point masses.

An advantage of the GLQ approach is that the accuracy of integration can be
controlled by the number of point masses used.

A disadvantage is the increased computation time as the number of
point masses increases.

Thus, there is a trade-off between accuracy and computation time, a common
theme in numerical methods.

\citet{Wild-Pfeiffer2008} investigated the use of different mass elements,
including tesseroids, to compute the gravitational effects of topographic
masses.

The author concludes that using tesseroids with GLQ integration gives the best
results for near-zone computations.

However, the question of how to determine the optimal parameters for GLQ
integration remained open.

Previous work by \citet{Ku1977} investigated the use of the Gauss-Legendre
Quadrature in gravity forward modeling.

\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.

The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.

Based on this, \citet{Ku1977} proposed an empirical criteria that the distance
between adjacent point masses should be less than the distance to the computation point.

\citet{Asgharzadeh2007} proposed using this criteria for the GLQ integration of
the gravity gradient tensor of tesseroids.

To our knowledge, there has been no investigation if the empirical criteria of
\citet{Ku1977} is valid for gravity gradient components or for tesseroids.

There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.

\citet{Li2011} devised an algorithm to automatically enforce the criteria of
\citet{Ku1977}.

Their algorithm divides the tesseroid into smaller ones instead of increasing
the number of point masses per tesseroid.

This division effectively increases the total number of point masses
because the number of points masses per tesseroid is fixed.

The criterion used decide if a division is necessary is
that the minimum distance to the computation point must be larger than
the largest dimension of the tesseroid.

This division is repeated recursively until all tesseroids obey the criteria.

Then, GLQ integration is performed for each of the smaller tesseroids
using the specified number of point masses.

The advantage of this adaptive discretization over
increasing the number of points masses is that the
total total distribution of point masses will be greater
only close to the computation point.

This makes the adaptive discretization more computationally efficient.

\citet{Grombein2013} developed optimized formula for the gravitational fields
of tesseroids using Cartesian integral kernels.

The Cartesian kernels are faster to compute.

The Cartesian kernels don't have singularities in the poles.

The Cartesian formula are numerically integrated using a Taylor series
expansion as proposed by \citet{Heck2007}.

\citet{Grombein2013} use a near-zone separation to mitigate the increased error
at high latitudes.

In the so called "near-zone" of the computation point they use a finner
discretization with smaller tesseroids.

This is accomplished by dividing the tesseroids along the horizontal
dimensions.

The algorithm of \citet{Li2011} is similar to the near-zone separation used by
\citet{Grombein2013}.

The determination of an optimal size of the near-zone remains an open question
\citep{Grombein2013}.

We have implemented a modified version of the adaptive discretion of
\citet{Li2011} into the open-source software package \textit{Tesseroids}.

The software uses the Cartesian formula of \citet{Grombein2013} for improved
performance and robustness.

Previous versions of \textit{Tesseroids} have been used by, e.g.,
\citet{Alvarez2012, Bouman2013, Bouman2013a, Mariani2013, Braitenberg2014,
Braitenberg2011, Fullea2014}.

This article describes the software \textit{Tesseroids} and the implementation
of our modified adaptive discretization algorithm.

We also present a numerical investigation of the error committed
in the computations.

These results allow us calibrate the adaptive discretization algorithm
separately for the gravitational potential, gravitational acceleration,
as well as the gravity gradient tensor components.

>>>>

Satellite missions dedicated to measuring the Earths gravity field
(like CHAMP, GRACE, and GOCE)
have provided geophysicists with almost uniform and global data coverage.
These new data have enabled interpretations on regional and global scales
\citep[e.g.][]{Reguzzoni2013,Braitenberg2015}.
Modeling at such scales requires taking into account the curvature of the
Earth and calculating gravity gradients as well as the traditional
gravitational acceleration.
A common approach to achieve this is
to discretize the Earth into tesseroids (Figure~\ref{fig:tesseroid})
instead of rectangular prisms.
However, the integral formula for the gravitational effects of a tesseroid
must be solved numerically.
Approaches to this numerical integration include
Taylor series expansion \citep{Heck2007, Grombein2013}
and the Gauss-Legendre Quadrature \citep{Asgharzadeh2007}.
Taylor series expansion produces accurate results at low latitudes but
presents a decrease in accuracy towards the polar regions.
This is attributed to tesseroids degenerating into an approximately triangular
shape at the poles.
The Gauss-Legendre Quadrature (GLQ) integration consists of approximating
the volume integral by a weighted sum of the effect of point masses.
An advantage of the GLQ approach is that the accuracy of integration can be
controlled by the number of point masses used.
A disadvantage is the increased computation time as the number of
point masses increases.
Thus, there is a trade-off between accuracy and computation time, a common
theme in numerical methods.
\citet{Wild-Pfeiffer2008} investigated the use of different mass elements,
including tesseroids, to compute the gravitational effects of topographic
masses.
The author concludes that using tesseroids with GLQ integration gives the best
results for near-zone computations.
However, the question of how to determine the optimal parameters for GLQ
integration remained open.

Previous work by \citet{Ku1977} investigated the use of the Gauss-Legendre
Quadrature in gravity forward modeling.
\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.
The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.
Based on this, \citet{Ku1977} proposed an empirical criteria that the distance
between adjacent point masses should be less than the distance to the computation point.
\citet{Asgharzadeh2007} proposed using this criteria for the GLQ integration of
the gravity gradient tensor of tesseroids.
To our knowledge, there has been no investigation if the empirical criteria of
\citet{Ku1977} is valid for gravity gradient components or for tesseroids.
There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.

\citet{Li2011} devised an algorithm to automatically enforce the criteria of
\citet{Ku1977}.
Their algorithm divides the tesseroid into smaller ones instead of increasing
the number of point masses per tesseroid.
This division effectively increases the total number of point masses
because the number of points masses per tesseroid is fixed.
The criterion used decide if a division is necessary is
that the minimum distance to the computation point must be larger than
the largest dimension of the tesseroid.
This division is repeated recursively until all tesseroids obey the criteria.
Then, GLQ integration is performed for each of the smaller tesseroids
using the specified number of point masses.
The advantage of this adaptive discretization over
increasing the number of points masses is that the
total total distribution of point masses will be greater
only close to the computation point.
This makes the adaptive discretization more computationally efficient.

\citet{Grombein2013} developed optimized formula for the gravitational fields
of tesseroids using Cartesian integral kernels.
These formula are faster to compute and don't have singularities at the poles
like their spherical counterparts.
The Cartesian formula are numerically integrated using a Taylor series
expansion as per \citet{Heck2007}.
\citet{Grombein2013} use a near-zone separation to mitigate the increased error
at high latitudes.
In the so called ``near-zone'' of the computation point they use a finner
discretization with smaller tesseroids.
This is accomplished by dividing the tesseroids along their horizontal
dimensions.
However, the determination of an optimal size of the near-zone remains an
open question \citep{Grombein2013}.

We have implemented a modified version of the adaptive discretion of
\citet{Li2011} into the open-source software package \textit{Tesseroids}.
The software uses the Cartesian formula of \citet{Grombein2013} for improved
performance and robustness.
Previous versions of \textit{Tesseroids} have been used by, e.g.,
\citet{Alvarez2012, Bouman2013, Bouman2013a, Mariani2013, Braitenberg2014,
Braitenberg2011, Fullea2014}.
This article describes the software \textit{Tesseroids} and the implementation
of our modified adaptive discretization algorithm.
We also present a numerical investigation of the error committed
in the computations.
These results allow us calibrate the adaptive discretization algorithm
separately for the gravitational potential, gravitational acceleration,
as well as the gravity gradient tensor components.
